---
title: "NYC_GEOsegmenation"
output: html_document
date: "2023-08-28"
---

# Exploratory Data Analysis and Clustering Analysis of New York City Census Tracts

This Markdown document presents an exploratory data analysis (EDA) and clustering analysis conducted on demographic and geographic data of New York City (NYC) census tracts. The analysis utilizes the R programming language and several libraries to visualize and understand patterns.

## Overview

In this analysis, we explore demographic and population density data for selected NYC census tracts using the `tidycensus`, `tigris`, `tidyverse`, `sf`, and other R packages. Census tracts serve as the building blocks of our analysis, providing a finer level of geographic granularity than zip codes and enabling us to uncover insights about different neighborhoods based on features such as median household income, median age, and population density.

## Methodology

1.  **Installation and Loading of Libraries:** We begin by installing and loading necessary R libraries, such as `tidycensus` for fetching census data, `sf` for spatial data handling, and visualization libraries like `ggplot2` and `mapview`.

2.  **Data Preparation:** Demographic data including median income, median age, and total population is fetched from the US Census Bureau using the `tidycensus` package. We calculate population density and perform data cleaning to prepare it for analysis.

3.  **K-Means Clustering:** To identify natural clusters within the data, we use the K-means clustering algorithm. The elbow method is employed to determine the optimal number of clusters that best captures the underlying patterns in the dataset.

4.  **Data Visualization:** The clustered data is converted to a geographic format using `sf` and visualized using various packages like `ggplot2`. We create an interactive map to visualize the distribution of clusters across NYC census tracts. Additionally, a scatter plot is generated to explore relationships between key demographic variables within each cluster.

5.  **Exporting GeoJSON:** We save the clustered data in a GeoJSON file format using the `sf` package, enabling further spatial analysis and data integration.

## Insights and Applications

This analysis offers valuable insights into NYC census tracts demographics and neighborhood patterns. It leverages census tracts, which provide a more nuanced view of geographic divisions than zip codes. The insights gained from this analysis can be utilized for various purposes:

-   Real estate investment decisions can be informed by identifying neighborhoods with similar characteristics.
-   Businesses can target specific demographic clusters for site selection and marketing strategies.
-   Urban planners can gain insights into neighborhood change and social integration trends.
-   Social services agencies can strategically allocate resources based on identified clusters.

By employing census tracts as our geographic units of analysis, we can better capture the diversity and nuances of NYC neighborhoods, contributing to informed decision-making.

**Note:** The full R code for this analysis can be found within this Markdown document. It covers data acquisition, preparation, clustering, and visualization.

## Installation and Loading of Libraries

```{r}

install.packages(c("tidycensus", "tigris", "tidyverse", "mapview", "ggiraph", "patchwork","ggplot2"))
install.packages("sf", type = "binary")
library(tidycensus)
library(tigris)
library(tidyverse)
library(sf)
library(mapview)
library(ggiraph)
library(patchwork)
library(ggplot2)

```

## Data Preparation

```{r}
# Set options and seed
options(tigris_use_cache = TRUE)
set.seed(123456)

# Fetch NYC data
nyc_counties <- c("005", "047", "061", "081", "085")

nyc_inputs <- get_acs(
  geography = "tract",
  variables = c(median_income = "B19013_001",
                median_age = "B01002_001",
                total_population = "B01003_001"),
  state = "NY",
  county = nyc_counties,
  output = "wide",
  geometry = TRUE,
  keep_geo_vars = TRUE
) %>%
  mutate(pop_density = total_populationE / (ALAND / 2589988.11 )) %>%
  na.omit()

```

## Elbow Method for Determining Cluster Amount

```{r}
# Elbow method to find optimal number of clusters
wcss <- numeric(length = 10)
for (num_clusters in 1:10) {
  nyc_kmeans <- nyc_inputs %>%
    st_drop_geometry() %>%
    select(median_incomeE, median_ageE, pop_density) %>%
    scale() %>%
    kmeans(centers = num_clusters)
  
  # Calculate the within-cluster sum of squares
  wcss[num_clusters] <- nyc_kmeans$tot.withinss
}

# Plot the elbow curve
plot(1:10, wcss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters", ylab = "Total Within-Cluster Sum of Squares")

```

## K-Means Clustering

```{r}

# Choose the number of clusters based on the elbow plot
num_clusters_chosen <- 4
nyc_kmeans <- nyc_inputs %>%
  st_drop_geometry() %>%
  select(median_incomeE, median_ageE, pop_density) %>%
  scale() %>%
  kmeans(centers = num_clusters_chosen)

```

### Data Visualization

Once we have performed K-means clustering on the dataset, we can visualize the results to gain insights into the distribution of clusters across NYC census tracts.

#### Clustered Map

The following code generates an interactive map that visualizes the clusters using different colors:

```{r}

# Create a clustered map
nyc_clusters <- nyc_inputs %>%
  mutate(cluster = as.character(nyc_kmeans$cluster))

nyc_map <- ggplot(nyc_clusters, aes(fill = cluster, data_id = GEOID)) + 
  geom_sf_interactive(size = 0.1) + 
  scale_fill_brewer(palette = "Set1") + 
  theme_void() + 
  labs(fill = "Cluster ")

# Display the clustered map
nyc_map
```

### Scatter Plot

Additionally, we can create a scatter plot to understand the relationships between median household income and population density for each cluster:

```{r}
# Create a scatter plot
nyc_plot <- ggplot(nyc_clusters, 
                   aes(x = median_incomeE, y = pop_density, color = cluster, data_id = GEOID)) + 
  geom_point_interactive() + 
  scale_color_brewer(palette = "Set1") + 
  scale_y_log10() + 
  scale_x_continuous(labels = scales::dollar_format()) + 
  theme_minimal(base_size = 12) + 
  labs(color = "Cluster",
       x = "Median household income",
       y = "Population density (logged)")

# Display the scatter plot
nyc_plot

```

## Exporting GeoJSON

We can also export the clustered data to a GeoJSON file, which allows for further spatial analysis:

```{r}
# Convert the data to a simple feature (sf) object
nyc_sf <- st_as_sf(nyc_clusters)

# Define the file path where you want to save the GeoJSON on your desktop
output_geojson <- "YOUR_FILE_PATH_HERE.geojson"

# Add cluster as an attribute to the sf object
nyc_sf$cluster <- nyc_clusters$cluster

# Write the sf object to a GeoJSON file, including the cluster attribute
st_write(nyc_sf, output_geojson)


```
